pdp: An R Package for Constructing Partial Dependence Plots

### 2017-01-01 manuscript/reviews/response.Rmd

### Comment 1.1
**Bug**: A graph is produced as expected if the code below is cut-and-pasted at the command line .  However, if the code is put into a file (say `foo.R`) and `source("foo.R")` is executed at the command line, no graph is produced.  (But note that if `plot=TRUE` is added to the call to `partial()`, then a plot is produced as expected.)
```{r, eval=FALSE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
library(pdp)
pd <- partial(ozone.rf, pred.var="Temp")
plotPartial(pd)
```

### Response 1.1
This is simply a side effect of using `lattice` to produce the plots. If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))` or `print(partial(fit, pred.var = "x", plot = TRUE))`). Additional text has been added to manuscript (see **Note:** on page 3) and the package documentation explaining this behavior. The fact that this worked when calling `partial` with `plot = TRUE` was a bug (which is now fixed).

### Comment 1.2
I couldn't get `partial` to work with `gbm` (version 2.1.1) models (although I didn't try very hard).  Perhaps this is because the `gbm` maintainer changed the `gbm` interface in a non-backcompatible way. The following (where `gbm.mod` is a model built using the gbm package)
```{r, eval=FALSE}
partial(gbm.mod, pred.var="age")
```
gives
```{r, eval=FALSE}
Error in paste("Using", n.trees, "trees...\n") :
argument "n.trees" is missing, with no default
```
I also tried
```{r, eval=FALSE}
partial(gbm.mod, pred.var="age", n.trees=100)
```
but that gives
```{r, eval=FALSE}
Error in .fun(piece, ...) : unused argument (n.trees = 100)
```

### Response 1.2
This has been fixed in `pdp` (version 0.3.0). The user may still have to supply `n.trees` in the call to `partial` (depending on which version of `gbm` they are using). Also, as suggested in another comment, a `/slowtests` directory is underway and will help prevent future issues like this. Thanks for cathing this.

@@

### Response 1.15
This is a good suggestion and is currently under development: https://github.com/bgreenwell/pdp/blob/master/slowtests/slowtests.R. In fact, the `gbm` maintainers have already contacted me about a breaking change caused by the next planned release of their package.

# -----------Reviewer 2-----------

### Comment
The Introduction immediately introduces the PDP. You must first motivate why it is important. In a linear model, they are not important. In a linear model with simple interactions, likewise. They first become important if you have non-parametric machine-learning techniques. Further, they are not the only idea extant when trying to explain a predictor's impact on the response. This paper would benefit if you can give a bit more of a lit review here.

### Response
I agree that PDPs are not as useful in "simple" linear models (e.g., few variables, only two-way interactions, etc.), but certainly there will be cases where PDPs will be extremely useful (e.g., many terms, complex interactions, etc.). In fact, linear models are still common in machine learning/computer science/data science (they just typically contain many terms). I also agree that PDPs are not the only, or sometimes best, approach to understanding the impact of a subset of the predictor's on the response. Brief descriptions and references are given for many alternatives: marginal model plots (`plotmo`), individual conditional expectation plots (`ICEbox`)---which are described as an alternatve to PDPs when strong interactions are suspected, effect plots (`effects`), and a few other techniques available in the `car` package. Unless the editor disagrees, I don't think it is necessary to go into more detail than what is already provided at the end of the introduction. Nonetheless, I have expanded the introduction (which links to the Boston housing example used throughout the paper) and included more background on the `ICEbox` package.

@@

### Comment
All `pdp` figures throughout the paper: Can you make explicit in the caption what code produces the left and right subfigures? I would use the subfigure package in latex and then put a comment in the R code e.g. # Fig 3b.

### Response
Rather than using subfigures, which I think would look messy with all the code, I made quite an attempt to comment which pieces of code produce which figure. Also, in response to another reviewer's comments, the code which produces figures containing multiple plots is better described and demonstrated. See, for example, the section called Multi-predictor PDPs.

### Edits:

### Comment
p1 abstract: strict assumptions imposed by traditional statistical models

### Response
Added some clarification here; in particular, normality, constant variance, and linearity.

@@

### Comment 2.9
Fig 7: as beautiful as this is, you should label the sidebar as "logit" or "centered logit" instead of "y".

### Response 2.9
Fixed. Color key labeled "Centered logit".

### Suggestions for the software

### Comment
I understand that you like the feature of having a one-shot function partial to both generate and plot the PDP. I realize that the default is `FALSE`. However, I still think it's a disservice to the user as constructing PDP's is computationally intense and if they get the plot wrong or it's not to their liking, they will have to regenerate the whole thing again. So, it is my opinion that you should take it out and force two steps: generation and illustration.

### Response
While I am not suggesting that the `partial` function needs to mimic the behavior of the PDP fucntions in `randomForest` and `gbm`, I do find it useful to be able to plot the results without forcing the user to store them. However, I agree with your concern which is why `partial` has `plot = FALSE` as the default. Furthermore, I have made it possible to retrieve the raw PDP data even when \code{plot = TRUE}. This is because the "trellis" object returned by `partial` \ `plotPartial` now contains an additional `partial.data` attribute. This is described in the package documentation. For example,
```{r, eval=TRUE}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)
pdp.Girth <- lattice::trellis.last.object()
pd.Girth <- attr(pdp.Girth, "partial.data")
head(pd.Girth)
```

### Comment
The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use random forests to illustrate classification probabilities.

### Response
This is a good suggestion. First, I have changed the behavior of `...` to refer to additional arguments in `predict` (e.g., the `n.trees` argument in `predict.gbm`.). Second, `partial` gained a new argument, `pred.fun`, where the user can provide an optional prediction function. This makes `partial` far more flexible. For example, you can use this argument to construct PDPs on the probability scale, or to construct ICE curves. These are illusrated in the new section labeled: **User-defined prediction functions**

### 2017-01-01 manuscript/greenwell.tex
minor changes

PDPs are essentially just averaged predictions; see, for example, step 1. (c) in Algorithm~\ref{alg:pdp}. As pointed out by \citet{goldstein-peeking-2015}, since traditional PDPs are based on the sample mean, strong interactions can conceal the complexity of the modeled relationship between the response and predictors of interest. This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE curves.

With \code{partial} it is possible to replace the mean in step 1. (c) with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale. It is even possible to obtain ICE curves. This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0). This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}. The supplied prediction function must return either a single prediction or a vector of predictions. If the mean of all the predictions is returned, the result will be equivalent to the traditional PDP. If a vector of predictions is returned (i.e., one for each observation), the result will be equivalent to the ICE plot procedure. The examples below illustrate.

Using the \code{pred.fun} argument it is possible to obtain PDPs for classification problems on the probability scale; we just need to write a function that extracts the predicted class probabilities of interest and then averages the result. The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
\begin{example}
pred.prob <- function(object, newdata) {  # see ?predict.svm
  pred <- predict(object, newdata, probability = TRUE)
  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
  mean(prob.setosa)
}
\end{example}
Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. The following chunk of code uses \code{partial} and \code{grid.arrange} to construct PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale. The results are displayed in Figure~\ref{fig:partial_iris_prob}.
\begin{example}
# Display plots side by side (Figure 8)
grid.arrange(
  partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob, plot = TRUE),
  partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob, plot = TRUE),
  partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
          pred.fun = pred.prob, plot = TRUE),
  ncol = 3
)
\end{example}

@@

Since the PDP is just the average of the corresponding ICE curves, it is quite simple to display both on the same plot. This can be accomplished by using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together. The code snippet below uses \pkg{ggplot2} and \code{grid.arrange} to plot ICE curves and c-ICE curves for the predictor \code{rm} in the Boston housing example. The results are displayed in Figure~\ref{fig:ice_cice_boston}.
\begin{example}
# ICE curves with their average
p1 <- ggplot(rm.ice, aes(rm, yhat)) +
  geom_line(aes(group = yhat.id), alpha = 0.2) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)

@@

# Display plots side by side (Figure 10)
grid.arrange(p1, p2, ncol = 2)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{ice_cice_boston}
  \caption{ICE curves (black curves) and their average (red curve) depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example. \textit{Left}: Uncentered (here the red curve is just the traditional PDP). \textit{Right}: Centered.}
  \label{fig:ice_cice_boston}
\end{figure}

@@

For illustration, we return to the Boston housing example. The code chunk below uses \pkg{caret} to tune an \pkg{xgboost} model using 10-fold cross-validation. (After loading \pkg{caret}, use \code{getModelInfo("xgbTree")} for information on tuning \pkg{xgboost} models.) \strong{Warning:} The following code chunk may take a few minutes to run.
\begin{example}
# Tune an XGBoost model using 10-fold cross-validation
library(caret)  # functions related to classification and regression training
set.seed(202)  # for reproducibility
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv, method = "xgbTree", metric = "Rsquared",
                    trControl = trainControl(method = "cv", number = 10),
                    tuneLength = 10)
\end{example}
The optimal model had a cross-validated $R^2$ of $0.902$ which was achieved using \code{nrounds = 100}, \code{max\_depth = 5}, \code{eta = 0.3}, \code{gamma = 0}, \code{colsample\_bytree = 0.8}, \code{min\_child\_weight = 1}, and \code{subsample = 0.9444444}. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
# Display plots side by side (Figure 11)
grid.arrange(
  partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
  ncol = 3
)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_xgb}
  \caption{PDPs for the top two most important variables in the Boston housing data using \pkg{xgboost}. Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
  \label{fig:boston_xgb}
\end{figure}

Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object there is no need to supply it in the call to \code{partial}. However, this is not the case when using the \pkg{xgboost} package directly. For example, the following code fits the same model using the \code{xgboost} function.
\begin{example}
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
                      colsample_bytree = 0.8, min_child_weight = 1,
                      subsample = 0.9444444)
\end{example}
To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}

>>>>

PDPs are essentially just averaged predictions; see, for example, step 1. (c) in Algorithm~\ref{alg:pdp}. Consequently, as pointed out by \citet{goldstein-peeking-2015}, strong interactions can conceal the complexity of the modeled relationship between the response and predictors of interest. This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE plot procedure.

With \code{partial} it is possible to replace the mean in step 1. (c) of Algorithm~\ref{alg:pdp} with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale. It is even possible to obtain ICE curves. This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0). This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}. The supplied prediction function must return either a single prediction or a vector of predictions. Returing the mean of all the predictions will result in the traditional PDP. Returing a vector of predictions (i.e., one for each observation) will result in a set of ICE curves. The examples below illustrate.

Using the \code{pred.fun} argument, it is possible to obtain PDPs for classification problems on the probability scale. We just need to write a function that computes the predicted class probability of interest averaged across all observations. The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
\begin{example}
pred.prob <- function(object, newdata) {  # see ?predict.svm
  pred <- predict(object, newdata, probability = TRUE)
  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
  mean(prob.setosa)
}
\end{example}
Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. The following chunk of code construct PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale. The results are displayed in Figure~\ref{fig:partial_iris_prob}.
\begin{example}
# PDPs for Petal.Width and Petal.Length on the probability scale
pdp.pw <- partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob,
                  plot = TRUE)
pdp.pl <- partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob,
                  plot = TRUE)
pdp.pw.pl <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                     pred.fun = pred.prob, plot = TRUE)

# Figure 8
grid.arrange(pdp.pw, pdp.pl, pdp.pw.pl, ncol = 3)
\end{example}

@@

Since the PDP is just the average of the corresponding ICE curves, it is quite simple to display both on the same plot. This can be accomplished by using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together. The code snippet below plots the ICE curves and c-ICE curves, along with their averages, for the predictor \code{rm} in the Boston housing example. The results are displayed in Figure~\ref{fig:ice_cice_boston}.
\begin{example}
# ICE curves with their average
p1 <- ggplot(rm.ice, aes(rm, yhat)) +
  geom_line(aes(group = yhat.id), alpha = 0.2) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)

@@

# Figure 10
grid.arrange(p1, p2, ncol = 2)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{ice_cice_boston}
  \caption{ICE curves (black curves) and their average (red curve) depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example. \textit{Left}: Uncentered (here the red curve is just the traditional PDP). \textit{Right}: Centered.}
  \label{fig:ice_cice_boston}
\end{figure}

@@

For illustration, we return to the Boston housing example. The code chunk below uses \pkg{caret} to tune an \pkg{xgboost} model using 10-fold cross-validation. (After loading \pkg{caret}, use \code{getModelInfo("xgbTree")} for information on tuning \pkg{xgboost} models.) \strong{Warning:} The following code chunk may take a few minutes to run.
\begin{example}
# Tune an XGBoost model using 10-fold cross-validation
library(caret)  # functions related to classification and regression training
set.seed(202)  # for reproducibility
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv, method = "xgbTree", metric = "Rsquared",
                    trControl = trainControl(method = "cv", number = 10),
                    tuneLength = 10)
\end{example}
The optimal model had a cross-validated $R^2$ of $0.902$ (use \code{print(boston.xgb\$bestTune)} to view the optimum tuning parameters). The next snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
# PDPs for lstat and rm
pdp.lstat <- partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE)
pdp.rm <- partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE)
pdp.lstat.rm <- partial(boston.xgb, pred.var = c("lstat", "rm"),
                        plot = TRUE, chull = TRUE)

# Figure 11
grid.arrange(pdp.lstat, pdp.rm, pdp.lstat.rm, ncol = 3)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_xgb}
  \caption{PDPs for the top two most important variables in the Boston housing data using \pkg{xgboost}. Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
  \label{fig:boston_xgb}
\end{figure}
%\code{nrounds = 100}, \code{max\_depth = 5}, \code{eta = 0.3}, \code{gamma = 0}, \code{colsample\_bytree = 0.8}, \code{min\_child\_weight = 1}, and \code{subsample = 0.9444444}

The \code{train} function creates objects of class \code{"train"}, whereas the \code{xgboost} function creates objects of class \code{"xgb.Booster"}. Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object, there is no need to supply it in the call to \code{partial} in this example. However, this is not the case when using the \pkg{xgboost} package directly. To illustrate, we fit the same model using the \code{xgboost} function with the tuning parameters found previously using \pkg{caret}.
\begin{example}
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
                      colsample_bytree = 0.8, min_child_weight = 1,
                      subsample = 0.9444444)
\end{example}
To use \code{partial} with \code{"xgb.Booster"} objects, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}
